Handbook streamFind: back end

### 2022-12-02 handbooks/Handbook_backend_streamFind.Rmd
LFU27 - updated basic processing, readme, handbook back end

The streamFind package is a back end (code based) and frontend ([shiny](https://shiny.rstudio.com/) app) platform for assembly of modular workflows to support spectrometric and spectroscopic data processing. The major focus of streamFind is data processing for environmental and quality studies. The streamFind package aims to stimulate the use of advanced data analysis (e.g., non-target screening, statistical analysis, etc.) in routine studies, promoting standardization of data processing and structure and easing the retrospective evaluation of data. The streamFind package can be used by academics but also by technicians due to the comprehensive documentation and well categorized set of integrated functionalities (modules). This handbook focuses on describing and demonstrating the back end of the streamFind package.

@@

For installation of the streamFind package, it is recommended to firstly ensure that the dependencies are currently installed. Several existing R packages and external functionalities are used in streamFind for various processing steps. A major dependency of streamFind is the [patRoon](https://github.com/rickhelmus/patRoon) R package.`r # TODO add reference` The patRoon package combines several functionalities for basic and advanced data processing and can be used interchangeably with streamFind, see \@ref(patroon-interchangeability) for more information. Most of the tools that patRoon depends on are equally used in streamFind. In the next sub-chapter, the mandatory and optional dependencies are described with installation instructions. For a standalone and self-contained usage option, the streamFind docker image is recommended (see \@ref(docker-image)), avoiding the need to installing all required dependencies.

## Install dependencies

### R and RTools

[R](https://cran.r-project.org/) and [RTools](https://cran.r-project.org/bin/windows/Rtools/) are essential dependencies. The `r R.version.string` can be obtained in <https://cran.r-project.org/>. The RTools can be downloaded in <https://cran.r-project.org/bin/windows/Rtools/>. Make sure to download the right version for the R version installed. Installation instructions are given in both sources.

### Third party software

The external third party dependencies recommended are listed in the section [other dependencies](https://rickhelmus.github.io/patRoon/handbook_bd/manual-installation.html#other-dependencies) from the patRoon [handbook](https://rickhelmus.github.io/patRoon/handbook_bd/manual-installation.html#r-prerequisites). Below we list the mandatory and really recommended software and resources.

-   [ProteoWizard](https://proteowizard.sourceforge.io/), required for conversion of vendor MS formats to mzML/mzXML;\
-   [Java JDK](https://www.oracle.com/java/technologies/downloads/), mandatory for patRoon when plotting structure using MetFrag;\
-   [OpenMS](https://www.openms.de/downloads/); esternal tool for peak picking, grouping and alignment;\
-   [MetFrag CL](http://ipb-halle.github.io/MetFrag/projects/metfragcl/), used for formula annotation in patRoon;\
-   [PubChemLite](https://zenodo.org/record/6503754#.YtqxE3bP1hE), database used with MetFrag in patRoon;\
-   [SIRIUS](https://bio.informatik.uni-jena.de/software/sirius/), tool for peak picking and formula annotation in patRoon;\
-   [BioTransformer](https://bitbucket.org/djoumbou/biotransformer/downloads/), used for prediction of transformation products in patRoon;\
-   [SAFD](https://bitbucket.org/SSamanipour/safd.jl/src/master/), tool for peak picking with data in profile mode;\
-   [Open Babel](http://openbabel.org/wiki/Main_Page), used in patRoon for validation of chemicals properties (e.g., InChI and formulae);

### R packages

The following code lines are from the patRoon [handbook](https://rickhelmus.github.io/patRoon/handbook_bd/manual-installation.html#r-prerequisites) to ensure that both patRoon and other mutual R dependencies are installed properly.

```{r r-packages, eval=FALSE}
install.packages(c("BiocManager", "remotes"))

BiocManager::install(c("mzR", "xcms","CAMERA"))

remotes::install_github("blosloos/enviPick")

BiocManager::install("ropls")
remotes::install_github("rickhelmus/KPIC2")

BiocManager::install("InterpretMSSpectrum")
remotes::install_github("cbroeckl/RAMClustR")

remotes::install_github("blosloos/nontargetData")
remotes::install_github("blosloos/nontarget")

remotes::install_github("rickhelmus/cliqueMS")

# only needed for Bruker DataAnalysis integration
remotes::install_github("BSchamberger/RDCOMClient")

BiocManager::install(c("BiocStyle", "Rgraphviz")) 
remotes::install_github("KelseyChetnik/MetaClean")

remotes::install_github("rickhelmus/patRoon", upgrade = "never", dependencies = TRUE)
```

After installation of the patRoon R package, the location of each third party software should be given in an `.Rprofile` file as suggested in section [other dependencies](https://rickhelmus.github.io/patRoon/handbook_bd/manual-installation.html#other-dependencies) of the patRoon [handbook](https://rickhelmus.github.io/patRoon/handbook_bd/manual-installation.html#r-prerequisites). For facilitation the same code lines are given below. The function `patRoon::verifyDependencies()` from patRoon can be used to verify if all dependencies are found.

```{r options-third-party-software, eval=FALSE}
options(patRoon.path.pwiz = "C:/ProteoWizard") # location of ProteoWizard installation folder
options(patRoon.path.SIRIUS = "C:/sirius-win64-3.5.1") # location where SIRIUS was extracted
options(patRoon.path.OpenMS = "/usr/local/bin") # directory with the OpenMS binaries
options(patRoon.path.pngquant = "~/pngquant") # directory containing pngquant binary
options(patRoon.path.MetFragCL = "~/MetFragCommandLine-2.4.8.jar") # full location to the jar file
options(patRoon.path.MetFragCompTox = "C:/CompTox_17March2019_SelectMetaData.csv") # full location to desired CompTox CSV file
options(patRoon.path.MetFragPubChemLite = "~/PubChemLite_exposomics_20220429.csv") # full location to desired PubChemLite CSV file
options(patRoon.path.BioTransformer = "~/biotransformer/biotransformer-3.0.1.jar")
options(patRoon.path.obabel = "C:/Program Files/OpenBabel-3.0.0") # directory with Open Babel binaries
```

## Install streamFind

The streamFind package can be installed from the GitHub repository <https://github.com/ricardobachertdacunha/streamFind> with the following code line.

@@

To be done.

@@

The streamFind package includes example data files via the streamFindData R package. Once installed the full file paths of MS data files can be obtained as shown below. Alternatively, the files can be obtained directly from folder `system.file(package = "streamFindData", dir = "extdata")`. The files available are described in Table \@ref(tab:description-files). Note, the MS files (.mzML) were trimmed with the function `trimSpectraFilesMZR` to reduce their size. The centroided Agilent Q-TOF files were trimmed with retention time between 900 and 1350 seconds, *m/z* for MS1 data between 200 and 450 Da and *m/z* for MS2 between 35 and 450 Da. Additionally, MS1 and MS2 traces below 100 and 50 counts were removed, respectively. The profile Agilent Q-TOF files were trimmed with retention time between 1000 and 1200 seconds, *m/z* for MS1 data between 200 and 300 Da and *m/z* for MS2 between 35 and 300 Da. Additionally, MS1 and MS2 traces below 1 count were removed to exclude empty spectra. The files are used within the document for demonstration of the back end.

```{r resource-files, cache=FALSE}
# 
# r_path <- system.file(package = "streamFindData", dir = "extdata")
# #r_path <- system.file(package = "streamFind", dir = "extdata")
# 
# files <- list.files(r_path, pattern = ".mzML|.mzXML", full.names = TRUE)

#list of file full paths from the data package
files <- streamFindData::msFilePaths()
```

@@

The streamFind package uses S4 class objects (e.g., *streamSet*, *msData*, *msAnalysis*, etc.) for structuring data. The S4 classes and their main getter and setter methods are presented and exemplified in this section. An overview of the S4 classes and their hierarchy is presented in Figure \@ref(fig:classes-overview) using MS analyses as an example. Other data types, such as from UV and Raman analyses, would lead to different classes (e.g., *uvData* and *uvAnalysis* and *ramanData* and *ramanAnalysis*, respectively) but the hierarchy is similar as shown for MS data.

@@

## Nomenclature

@@

The set creation is demonstrated below for the three MS files (triplicate, 1 to 3 in \@ref(tab:description-files)) where a set of chemicals were spiked (see \@ref(spiked-chemicals) with tag IS and S). The set path can be assigned as argument but for the example below it was left as default (i.e., the working directory). The initial S4 class created in the `newStreamSet` function is *streamSet*. The subclass is then defined according to the type of files added. Note that multiple file types are not possible as a subclass cannot be defined.

```{r newstreamset-call, message=FALSE, results='hide'}
#New streamSet with the files 1 to 3
sp1 <- newStreamSet(files = files[1:3], title = "Example_01")
```

@@

Because the added files were all with mzML format (i.e., MS files), the subclass is directly defined as *msData*. The structure for further processing data is then set for MS data. Other data formats/structure will lead to different sub classes (e.g., *uvData* and *ramanData*) but is not yet implemented within the streamFind package.

@@

The *msData* structure includes the *streamSet* class plus the `features` slot. The title, date and path slots correspond to the set title, creation date and directory in disk, respectively. The analyses slot is a list with a *msAnalysis* S4 class object per MS file added. The structure of the *msAnalysis* class is described in the next sub-chapter. The `features` slot is a *msFeatures* S4 class object to hold the results from alignment, grouping and annotation of peaks across analyses. Further information for *msAnalysis* is given in the next sub-chapter (\@ref(msanalysis-class)) and for *msFeatures* is given in the sub-chapter \@ref(msfeatures-class), after creation of features by performing alignment and grouping of peaks across analyses.

@@

The slot analysis corresponds to the analysis name. The metadata slot is a list of objects with informative entries of the analysis file. The metadata list can be expanded with additional information about the analysis/sample (e.g., sample location, weather conditions, process sensor data, etc.). The slot parameters is used to store processing steps (e.g., peak picking) as *settings* S4 class objects, including the processing step name/ID, the algorithm used and the respective settings applied. The parameters can be used for history track but also for re-run the processing steps when required. The spectra and chromatograms slots are used to store low level data (i.e., mass traces either in profile or centroided modes) from the MS files. Finally, the peaks slot is used to store the chromatographic peaks after peak picking. As no other processing step was applied the spectra, chromatograms and peaks tables in the *msAnalysis* are empty for now.

@@

As aforementioned, access to raw data is based on defining *m/z* and retention time targets. The function `makeTargets` is used for comprehensively build targets. As shown below for Carbamazepin d10 and Diuron d6, multiple ways can be used to assemble targets but similarly returning a uniform `data.table` class object with a target list. Accessing and plotting S4 methods in streamFind use the `makeTargets` function to collect raw data from the analysis files. Therefore, the arguments `mz`, `rt`, `ppm`, `sec` and `id` are present in most S4 methods for accessing and plotting data as demonstrated in the following sections.

@@

## TICs

Total ion chromatograms (TICs) are obtained either by pre-loading and adding the chromatogram data to the *msAnalysis*, extracting the data in a `TICs` call (S4 method) or loading all spectra and applying the sum of intensities to each spectrum. Note, TIC is already available as chromatogram in mzML data from the major HRMS vendors. The S4 method TICs first evaluates if the TIC chromatogram is available in the mzML and when not, the spectra data is extracted and used instead.

```{r tics_method}
# TIC for both analyses in the msData class object
tic1 <- TICs(sp1, analyses = 1:3)
tic1
```

```{r tics_method-2}
#TIC of the msAnalysis object
head(TICs(a1), 6)
```

Similarly to EICs, the TICs can be plotted with `plotTICs` S4 method, as shown below.

@@

@@

```{r loadrawdata-msanalysis}
#load raw data to an msAnalysis object, while setting a minimum intensity value for both MS1 and MS2 traces
a1 <- loadRawData(a1, minIntensityMS1 = 0, minIntensityMS2 = 100)
```

@@

```{r loadrawdata-msdata}
#load raw data to the first three samples in an msData object.
sp1 <- loadRawData(sp1, analyses = 1:3)
```

@@

@@

@@

`r # TODO create method to access chromatograms in msAnalysis and msData`

@@

Regarding MS data, the format used in streamFind is .mzML or .mzXML either centroided or profile. However, both mentioned formats do not correspond to the formats of actual MS system vendors. [msConvert](https://proteowizard.sourceforge.io/download.html) from [ProteoWizard](https://proteowizard.sourceforge.io/) can be used to convert the main MS formats (and others) to .mzML or .mzXML, including data centroiding when required. The msConvert GUI tool can be used for conversion. However, if automation is required, the [command line](https://proteowizard.sourceforge.io/tools/msconvert.html) of msConvert is used within the function `convertFiles` in streamFind, as shown below. The function `compatibleFileFormatsForConversion()` can be used to check which vendor formats are currently possible to convert in streamFind through ProteoWizard.

@@

The analysis set creation is demonstrated below for twelve HRMS data files corresponding to blank and wastewater influent samples, both measured in triplicate and in positive and negative modes. The argument `files` is the main input and can either be a vector of full path files, a table with file (i.e., full file path), replicate (i.e., the name of the analysis replicate group for each file) and blank (i.e., the name of the associated blank analysis replicate group for each file), or the `analysisInfo` data.frame from the package [patRoon](https://github.com/rickhelmus/patRoon). The set path can be assign as argument `path` and is the directory where all the set files (i.e., scripts, cache databases, objects and results) will be stored by default.

```{r newstreamset-basicworkflow, message=FALSE, results='hide'}
sp2 <- newStreamSet(files = files[10:21], path = getwd(), title = "Project Example")
```

@@

Before preforming data processing, the analysis names (i.e., replicate name and associated blank replicate name) and metadata can be amended/extended.

@@

@@

The initial processing step of the basic workflow is the peak picking. Yet, other steps might be necessary prior the pick picking, such as data conversion, data calibration, etc. These are discussed later in section X. The basic workflow in streamFind is based on [patRoon](https://github.com/rickhelmus/patRoon) and most of the output in streamFind can be used/converted to enable the use of native [patRoon](https://github.com/rickhelmus/patRoon) functions and methods.

@@

```{r peakpicking-msdata-run, results='hide', message=FALSE}
#peak picking call using the stored parameters
sp2_pp <- peakPicking(sp2_pp)
```

@@

The processing parameters for grouping and alignment are added similarly to the processing settings for peak picking but the argument `where` is set to "features", as shown below. The settings are added to the slot parameters of the *msFeatures* object and can be obtained by the method `getParameters`, as shown below.

@@

The structure of the *msFeatures* is shown below. The slot `analyses` is the `data.table` as obtained by the `analysisTable`, containing basic information about the analyses, such as file full path, replicate and associated blank replicates. The intensity slot contains the intensity of the feature in each analysis (i.e., corresponding peak intensity in each analysis). The slot metadata holds additional information about each feature, such as average *m/z*/ neutral mass, retention time and number of corresponding peaks. The slot settings has the list of settings used to build/process features, such as peakGrouping and filterFeatures.

@@

```{r inspecting-features}
#getter for feature intensities correspondent to the mass of Diuron in each sample by setting average to false
features(sp2_pg, mass = diuron_d6, ppm = 10, sec = 20, average = FALSE)

@@

```{r grouping-msdata-only-positive}

sp2_pg_3 <- sp2_pp[c(4:6, 10:12)]

settings_pg_xcms_2 <- list(
  rtalign = TRUE,
  loadRawData = TRUE,
  groupParam = xcms::PeakDensityParam(
    sampleGroups = "holder",
    bw = 3,
    minFraction = 0.6,
    minSamples = 2,
    binSize = 0.008,
    maxFeatures = 100),
  preGroupParam = xcms::PeakDensityParam(
    sampleGroups = "holder",
    bw = 5,
    minFraction = 1,
    minSamples = 3,
    binSize = 0.008,
    maxFeatures = 100),
  retAlignParam = xcms::PeakGroupsParam(
    minFraction = 1,
    extraPeaks = 0,
    smooth = "loess",
    span = 0.3,
    family = "gaussian")
)

@@

sp2_pg_3 <- peakGrouping(sp2_pg_3, settings = settings_pg_xcms_2)

peaks(sp2_pg_3)

features(sp2_pg_3, mass = targets4, ppm = 10, sec = 10, complete = TRUE)

plotAlignment(sp2_pg_3)
```

@@

As peak picking can result in false negatives and to ensure that false positives are less likely, a recursive integration is often performed for filling missing peaks in a feature. The recursive integration consists of extracting the mass traces in the feature region from analyses not represented, returning the hipotetical peak intensity. The exact collection of mass traces and calculation of the peak intensity (or height) is dependent on the algorithm applied. Below, an example of the recursive integration workflow using the function [fillChromPeaks](https://rdrr.io/bioc/xcms/man/fillChromPeaks.html) from the package [xcms](https://bioconductor.org/packages/release/bioc/html/xcms.html) is shown. The function `peakFilling` is used for recursive integration and a *settings* S4 class object is given to define the algorithm and respective parameters. A *settings* object can be obtained with `fillingSettingsDefaultXCMS` for a default *ChromPeakAreaParam* from xcms. Note that the peakFilling settings are added to the parameters slot of the *msData* object.

@@

Annotation of isotopes and adducts is crucial not only to prioritize the relevant features but also to deliver essential information for further identification, such as charge, isotopic pattern and adduct yield.

@@

The results of annotation are stored under annotation of the *msFeatures* S4 class and matches the [components](https://rickhelmus.github.io/patRoon/reference/components-class.html) S4 class of the [patRoon](https://github.com/rickhelmus/patRoon) package. This means that the feature components can be accessed via the `getAnnotation` method for an *msData* object and inspected also with native methods from patRoon.

@@

Data filtration is crucial for proper prioritization of relevant peaks/features. In streamFind, two levels of data filtration is implemented: (1) data sub-setting (permanent) and (2) filter processing steps for peaks and/or features (tagged but not removed).

@@

@@

```{r quality-sn}

# sp2_pg <- calculateSNR(sp2_pg, targetsID = features(sp2_pg)[["id"]][1:5])
# 
# features(sp2_pg, complete = TRUE)[1:5, ]

@@

## Processing settings

Settings are needed to extract and average the fragmentation data. Below an example of settings used for accessing and averaging MS2 data is given. The resulting *settings* object is then used further in the call to `MS2s` or `plotMS2s`.

```{r create-settings-msn}
#the parameter settings to obtained and process MSn spectra
param_msn <- list(
  isolationTimeWindow = 0,
  isolationMassWindow = 1.3,
  clusteringMethod = "distance",
  clusteringUnit = "ppm",
  clusteringWindow = 10,
  minIntensityPre = 200,
  minIntensityPost = 300,
  asPatRoon = FALSE, #when TRUE, is only used by running generateMS2
  mergeVoltages = TRUE,
  mergeBy = "analyses"
)

#creating the settings S4 class
settings_msn <- createSettings(
  call = "extractMSn",
  algorithm = "streamFind",
  parameters = param_msn
)

# the class of settings 
is(settings_msn)
```

@@

The method `MS2s` can be used to collect MS2 data from given targets, following the same principle as `makeTargets` within `EICs`. Similarly, `plotMS2s` can be used to access and directly plot MS2 data from targets, as shown below. A more general function for accessing the data is `extractMSn`, where further fragmentation levels can be accessed. When using streamFind basic workflow, the use of `MS2s` and `plotMS2s` is recommended. All the S4 methods/function can be applied for *msAnalysis* and *msData* objects.

```{r extract-ms2}
# #head of MS2 data for target 1 in a msAnalysis object
# head(MS2s(a1, mz = targets4[1, ], settings = settings_msn))
# 
# #ploting the MS2 data of the first target in a msAnalysis object
# plotMS2s(a1, mz = targets4[1, ], settings = settings_msn, interactive = FALSE)
# 
# #most intense MS2 data traces for both targets in a msData object
# s_data_3 <- MS2s(sp1, mz = targets4, settings = settings_msn)
# s_data_3 <- s_data_3[order(intensity, decreasing = TRUE), ]
# head(s_data_3)
# 
# #ploting the MS2 data of the first target in a msAnalysis object
# plotMS2s(sp1, mz = targets4, settings = settings_msn, interactive = TRUE, colorBy = "targets")
```

>>>>

The streamFind project entitled "Flexible data analysis and workflow designer to identify chemicals in the water cycle" is funded by the [Bundesministerium für Bildung und Forschung (BMBF)](https://www.bmbf.de) and is a cooperation between the [Institut für Energie- und Umwelttechnik e. V. (IUTA)](https://www.iuta.de) and [Forschungszentrum Informatik (FZI)](https://www.fzi.de/). The major focus of streamFind is data processing for environmental and quality studies. The streamFind package aims to stimulate the use of advanced data analysis (e.g., non-target screening, statistical analysis, etc.) in routine studies, promoting standardization of data processing and structure and easing the retrospective evaluation of data. The streamFind package can be used by academics but also by technicians due to the comprehensive documentation and well categorized set of integrated functionalities or modules (see handbooks folder for more information).

This handbook focuses on describing and demonstrating the R package streamFind which aims to be a tool set for processing of MS, Raman, UV and other types of data.

@@

For installation of the streamFind package, it is recommended to firstly ensure that the dependencies are currently installed. Several existing R packages and external functionalities are used in streamFind for various processing steps. A major dependency of streamFind is the [patRoon](https://github.com/rickhelmus/patRoon) R package. The patRoon package combines several functionalities for basic and advanced data processing and can be used interchangeably with streamFind.

### Install dependencies

#### R and RTools

[R](https://cran.r-project.org/) and [RTools](https://cran.r-project.org/bin/windows/Rtools/) for Windows users. The `r R.version.string` can be obtained in https://cran.r-project.org/. The RTools can be downloaded in https://cran.r-project.org/bin/windows/Rtools/. Make sure to download the right version for the R version installed. Installation instructions are given in both sources.

#### patRoon

The patRoon's [handbook](https://rickhelmus.github.io/patRoon/handbook_bd/manual-installation.html#r-prerequisites) offers detailed information on how to install the [patRoon](https://github.com/rickhelmus/patRoon) R package and the required [dependencies](https://rickhelmus.github.io/patRoon/handbook_bd/manual-installation.html#other-dependencies).

### Install streamFind

The [streamFind](https://github.com/ricardobachertdacunha/streamFind) R package can be installed from the GitHub repository with the following code line:

@@

Not yet available.

@@

The streamFind package includes example data files via the [streamFindData R package](https://github.com/ricardobachertdacunha/streamFindData). Once installed the full file paths of MS data files can be obtained as shown below. Alternatively, the files can be obtained directly from folder `system.file(package = "streamFindData", dir = "extdata")`. The files available are described in Table \@ref(tab:description-files). Note, the MS files (.mzML) were trimmed with the function `trimSpectraFilesMZR` to reduce their size. The centroided Agilent Q-TOF files were trimmed with retention time between 900 and 1350 seconds, *m/z* for MS1 data between 200 and 450 Da and *m/z* for MS2 between 35 and 450 Da. Additionally, MS1 and MS2 traces below 100 and 50 counts were removed, respectively. The profile Agilent Q-TOF files were trimmed with retention time between 1000 and 1200 seconds, *m/z* for MS1 data between 200 and 300 Da and *m/z* for MS2 between 35 and 300 Da. Additionally, MS1 and MS2 traces below 1 count were removed to exclude empty spectra. The files are used within the document for demonstration of the back end.

```{r resource-files, cache=FALSE}
files <- streamFindData::msFilePaths()
```

@@

The streamFind package uses S4 class objects (e.g., *streamSet*, *msData*, *msAnalysis*, etc.) for structuring data. The S4 classes and their main getter and setter methods are presented and exemplified in this section. An overview of the S4 classes and their hierarchy is presented in Figure \@ref(fig:classes-overview) using MS analyses as an example. Other data types, such as from UV and Raman analyses, would lead to different classes (e.g., *uvData* and *uvAnalysis* and *ramanData* and *ramanAnalysis*, respectively) but the hierarchy is similar as shown for MS data (not yet implemented).

@@

@@

The set creation is demonstrated below for the three MS files (triplicate, 1 to 3 in \@ref(tab:description-files)) where a set of chemicals were spiked (see \@ref(spiked-chemicals) with tag IS and S). The initial S4 class created in the `newStreamSet` function is *streamSet*. The subclass is then defined according to the type of files added. Note that multiple file types are not possible as a subclass cannot be defined.

```{r newstreamset-call, message=FALSE, results='hide', message=FALSE, warning=FALSE}
#New streamSet with the files 1 to 3
sp1 <- newStreamSet(files = files[1:3], title = "Example_01")
```

@@

Because the added files were all with mzML format (i.e., MS files), the subclass of `sp1` is directly defined as *msData*. The structure for further processing data is then set for MS data. Other data formats/structure will lead to different sub classes (e.g., *uvData* and *ramanData*) but is not yet implemented within the streamFind package.

@@

The *msData* structure includes the *streamSet* class plus the `features` slot. The title and date slots correspond to the set title and creation date, respectively. The analyses slot is a list with a *msAnalysis* S4 class object per MS file added. The `features` slot is a *msFeatures* S4 class object to hold the results from alignment, grouping (or correspondence) and annotation of peaks across analyses. Further information for *msAnalysis* is given in the next sub-chapter (\@ref(msanalysis-class)) and for *msFeatures* is given in the sub-chapter \@ref(msfeatures-class), after creation of features by performing correspondence of peaks across analyses.

@@

The slot analysis corresponds to the analysis name. The metadata slot is a list of objects with informative entries of the analysis file. The metadata list can be expanded with additional information about the analysis/sample (e.g., sample location, weather conditions, process sensor data, etc.). The slot settings is used to store processing steps (e.g., peak picking) as *settings* S4 class objects, including the processing step name/ID, the algorithm used and the respective processing parameters applied. The settings can be used for history track but also for re-run the processing steps when required. The spectra and chromatograms slots are used to store low level data (i.e., mass traces either in profile or centroided modes) from the MS files. Finally, the peaks slot is used to store the chromatographic peaks after peak picking. As no other processing step was applied the spectra, chromatograms and peaks in the *msAnalysis* are empty `data.tables` for now.

@@

As aforementioned, access to raw data is based on defining *m/z* and retention time targets. The function `makeTargets` is used for comprehensively build targets. As shown below for Carbamazepin-d10 and Diuron-d6, multiple ways can be used to assemble targets but similarly returning a uniform `data.table` class object with a target list. Accessing and plotting S4 methods in streamFind use the `makeTargets` function to collect raw data from the analysis files. Therefore, the arguments `mz`, `rt`, `ppm`, `sec` and `id` are present in most S4 methods for accessing and plotting data as demonstrated in the following sections.

@@

## TICs and BPCs

Total ion chromatograms (TICs) and base peak chromatograms (BPCs) are obtained either by pre-loading and adding the chromatogram data to the *msAnalysis*, extracting the data in a `TICs` call (S4 method) or loading all spectra and applying the sum of intensities to each spectrum. Note, TIC is already available as chromatogram in mzML data from the major HRMS vendors. The S4 method TICs first evaluates if the TIC chromatogram is available in the mzML and when not, the spectra data is extracted and used instead.

```{r tics_method}
# TIC for both analyses in the msData class object
tic1 <- TICs(sp1, analyses = 1:3)
head(tic1, 6)
```

```{r tics_method-2}
#TIC of the msAnalysis object
head(TIC(a1), 6)
```

```{r bpc_method}
#BPC of the msAnalysis object
head(BPC(a1), 6)
```

Similarly to EICs, the TICs and BPCs can be plotted with `plotTICs` and `plotBPCs` S4 methods, respectively, as shown below.

@@

```{r plotbpcss-method-2}
#using the msData object
plotBPCs(sp1, colorBy = "replicates")
```

@@

```{r loadrawdata-msanalysis}
#load raw data to an msAnalysis object
a1 <- loadRawData(a1)
```

@@

```{r loadrawdata-msdata}
#load raw data to the first three samples in an msData object.
sp1 <- loadRawData(sp1)
```

@@

Spectra can be accessed from an *msAnalaysis* or *msData* as shown below with methods `spectra`. The method `plotSpectra` is also available for both classes.

@@

```{r plot-spectra-msanalysis}
# plotting spectra of isotopic pattern for Diuron-d6 in an msAnalysis
plotSpectra(a1, mz = data.frame(mzmin = 238, mzmax = 250,
                                rtmin = 1145, rtmax = 1170))
```

@@

Chromatograms can be included in mzML files. For instance, TIC is often available as chromatogram in mzML files. Also, analyses produced in multiple reaction monitoring (MRM) mode, when exported to mzML without collapsing the MRMs, will have the data stored as chromatograms, as shown below. The method `chromatograms`

```{r create-mrm-msAnalysis}
# MS files acquired in MRM
mrm <- newAnalysis(file = files[29])

# data only as chromatograms
mrm
```

```{r laod-show-chromatograms}
# loading raw data into the msAnalysis
mrm <- loadRawData(mrm)

# getting the chromatograms as data.table
chroms <- chromatograms(mrm)

head(chroms)
```

```{r plot-loaded-chromatograms}
# plotting the second and third chromatograms in the msAnalysis
plotChromatograms(mrm, index = c(2,3), interactive = TRUE)
```

@@

Regarding MS data, the format used in streamFind is mzML or mzXML either centroid or profile mode. However, both mentioned formats do not correspond to the formats of actual MS system vendors. [msConvert](https://proteowizard.sourceforge.io/download.html) from [ProteoWizard](https://proteowizard.sourceforge.io/) can be used to convert the main MS formats (and others) to mzML (preferable as it is the most recent) or mzXML, including data centroiding when required. The msConvert GUI tool can be used for conversion. However, if automation is required, the [command line](https://proteowizard.sourceforge.io/tools/msconvert.html) of msConvert is used within the function `convertFiles` in streamFind, as shown below. The function `compatibleFileFormatsForConversion()` can be used to check which vendor formats are currently possible to convert in the streamFind through ProteoWizard.

@@

The analysis set creation is demonstrated below for twelve HRMS data files corresponding to blank and wastewater (influent) samples, both measured in triplicate and in positive and negative modes. The argument `files` is the main input and can either be a vector of full path files, a table with file (i.e., full file path), replicate (i.e., the name of the analysis replicate group for each file) and blank (i.e., the name of the associated blank analysis replicate group for each file), or the `analysisInfo` data.frame from the package [patRoon](https://github.com/rickhelmus/patRoon). Setting `run_parallel` to `TRUE` will enable parallel processing, which can be useful for large number of files. See more details about parallel processing in the section \@ref(parallel-processing).

```{r newstreamset-basicworkflow, message=FALSE, results='hide', warning=FALSE}
sp2 <- newStreamSet(files = files[10:21],
                    title = "Project Example",
                    run_parallel = TRUE)
```

@@

Before preforming data processing, the analysis replicate names and associated blank replicate names as well as metadata can be amended/extended.

@@

`r # TODO add concatenation for msData objects`

@@

An initial processing step of the basic workflow is the peak picking. Other processing steps might be necessary prior the pick picking, such as data conversion, data centroiding, data calibration, etc. but are not yet implemented in streamFind. The peak picking and other basic processing steps in streamFind are applied through [patRoon](https://github.com/rickhelmus/patRoon). Both *msData* and *msAnalyses* can be converted to native [patRoon](https://github.com/rickhelmus/patRoon) objects for native use of [patRoon](https://github.com/rickhelmus/patRoon). See section \@ref(patRoon-interchangeability) for more information.

`r # TODO create as.msData and as.msAnalysis methods for patRoon's features/featureGroups`

@@

```{r peakpicking-msdata-run, results='hide', message=FALSE, warning=FALSE}
#peak picking call using the stored parameters
sp2_pp <- peakPicking(sp2_pp)
```

@@

The processing parameters for grouping and alignment are added similarly to the processing settings for peak picking but the argument `where` is set to "features", as shown below. The settings are added to the slot settings of the *msFeatures* object and can be obtained by the method `getSettings`, as shown below. Note that, for grouping positive and negative analyses, the retention time alignment would not work. Alignment can be applied only with either positive or negative analysis as shown below in \@ref(alignment-results).

@@

The structure of the *msFeatures* is shown below. The slot `analyses` is the `data.table` as obtained by the `analysisTable`, containing basic information about the analyses, such as file full path, replicate and associated blank replicates. The intensity slot contains the intensity of the feature in each analysis (i.e., corresponding peak intensity in each analysis). The slot metadata holds additional information about each feature, such as average *m/z*/ neutral mass, retention time and number of corresponding peaks. The slot settings has the list of settings used to build/process features, such as `peakGrouping` and `filterFeatures`.

@@

```{r inspecting-features, message=FALSE, warning=FALSE}
#getter for feature intensities correspondent to the mass of Diuron in each sample by setting average to false
features(sp2_pg, mass = diuron_d6, ppm = 10, sec = 20, average = FALSE)

@@

```{r grouping-msdata-only-positive, message=FALSE, warning=FALSE}

# subset of msData with only positive analyses
sp2_pg_3 <- sp2_pp[c(4:6, 10:12)]

# grouping settings with alignment scheduled 
settings_pg_xcms_2 <- list(
  rtalign = TRUE,
  loadRawData = TRUE,
  groupParam = xcms::PeakDensityParam(
    sampleGroups = "holder",
    bw = 3,
    minFraction = 0.6,
    minSamples = 2,
    binSize = 0.008,
    maxFeatures = 100),
  preGroupParam = xcms::PeakDensityParam(
    sampleGroups = "holder",
    bw = 5,
    minFraction = 1,
    minSamples = 3,
    binSize = 0.008,
    maxFeatures = 100),
  retAlignParam = xcms::PeakGroupsParam(
    minFraction = 1,
    extraPeaks = 0,
    smooth = "loess",
    span = 0.3,
    family = "gaussian")
)

@@

sp2_pg_3 <- peakGrouping(sp2_pg_3, settings = settings_pg_xcms_2)
```

@@

As peak picking can result in false negatives and to ensure that false positives are less likely, a recursive integration is often performed for filling missing peaks in a given feature. The recursive integration consists of extracting the mass traces in the feature region from analyses not represented, returning the hypothetical peak intensity. The exact collection of mass traces and calculation of the peak intensity (or height) is dependent on the algorithm applied. Below, an example of the recursive integration workflow using the function [fillChromPeaks](https://rdrr.io/bioc/xcms/man/fillChromPeaks.html) from the package [xcms](https://bioconductor.org/packages/release/bioc/html/xcms.html) is shown. The function `peakFilling` is used for recursive integration and a *settings* S4 class object is given to define the algorithm and respective parameters. A *settings* object can be obtained with `fillingSettingsDefaultXCMS` for a default *ChromPeakAreaParam* from xcms. Note that the `peakFilling` settings are added to the settings slot of the *msFeatures* object.

@@

Annotation of isotopes and adducts is crucial not only to prioritize the relevant features but also to deliver essential information for further identification, such as charge, isotopic pattern and adduct yield. Annotation in streamFind is applied through patRoon which uses the [CAMERA](https://www.bioconductor.org/packages/release/bioc/html/CAMERA.html) R package. Again, *msData* objects with both positive and negative analyses would not work for annotation. The results of annotation are stored as extra columns in the metadata slot of the *msFeatures* S4 class.

@@

@@

Data filtration is crucial for proper prioritization of relevant peaks/features. In streamFind, two levels of data filtration is implemented: (1) data sub-setting (permanent) and (2) filter processing steps for peaks and/or features (both are tagged as filtered but not removed).

@@

`r # TODO add filterPeaks option to msData and msAnalysis`

@@

`r # TODO implement MetaClean qualita evaluation``

```{r quality-sn}
sp2_pg <- calculateSNR(sp2_pg, targetsID = features(sp2_pg)[["id"]][1:2])

features(sp2_pg, complete = TRUE)[1:2, ]
```

@@

@@

The method `MS2s` can be used to collect MS2 data from given targets, following the same principle as `makeTargets` within `EICs`. Similarly, `plotMS2s` can be used to access and directly plot MS2 data from targets, as shown below.

```{r extract-ms2}
#head of MS2 data for target 1 in a msAnalysis object
head(MS2s(a1, mzClust = 0.01, isolationWindow = 1, mz = targets4[1, ]))

#plotting the MS2 data of the first target in a msAnalysis object
plotMS2s(a1, mzClust = 0.01, isolationWindow = 1, mz = targets4[1, ])

#most intense MS2 data traces for both targets in a msData object
ms2_data <- MS2s(sp1, mzClust = 0.01, isolationWindow = 1, mz = targets4)
ms2_data <- ms2_data[order(intensity, decreasing = TRUE), ]
head(ms2_data)

#plotting the MS2 data of the first target in a msAnalysis object
plotMS2s(sp1, mzClust = 0.01, isolationWindow = 1, mz = targets4,
         interactive = TRUE, colorBy = "targets")
```
